---
title: "Project Code"
format: html
editor: visual
---

### Goal: Determine if different states have different medicare payment amounts (money medicare provides), discern which factors contribute to this difference.

Tot_Mdcr_Stdzd_Amt (total medical medicare payment amount standrardized) is the response variable

### Libraries

```{r}
library(ggplot2)
library(dplyr)
```


### Lets load in the data

```{r}
phys_dat <- read.csv('Datasets/Medicare Physician & Other Practitioners - by Provider/2022/MUP_PHY_R24_P07_V10_D22_Prov.csv')
head(phys_dat)
```

```{r}
tax_dat <- read.csv('Datasets/21in55/21in55cmcsv.csv')
head(tax_dat)
```

### Data Cleaning

```{r}
# Let's only use providers in US states
phys_dat <- subset(phys_dat, Rndrng_Prvdr_State_Abrvtn %in% state.abb)

# Let's only use tax information from US states
tax_dat <- subset(tax_dat, STATE %in% state.abb)
```

```{r}
# Removing unnecessary variables

## phys_dat

phys_dat2 <- phys_dat

### name probably isn't important
phys_dat2$Rndrng_Prvdr_Last_Org_Name <- NULL
phys_dat2$Rndrng_Prvdr_First_Name <- NULL
phys_dat2$Rndrng_Prvdr_MI <- NULL

### We are not differentiating organizations from individual doctors
phys_dat2$Rndrng_Prvdr_Crdntls <- NULL
phys_dat2$Rndrng_Prvdr_Ent_Cd <- NULL
phys_dat2$Rndrng_Prvdr_Gndr <- NULL

### Non-state address info irrelevant
phys_dat2$Rndrng_Prvdr_St1 <- NULL
phys_dat2$Rndrng_Prvdr_St2 <- NULL
phys_dat2$Rndrng_Prvdr_City <- NULL
phys_dat2$Rndrng_Prvdr_State_FIPS <- NULL # Irrelevant with state abbreviations
phys_dat2$Rndrng_Prvdr_Zip5 <- NULL
phys_dat2$Rndrng_Prvdr_RUCA_Desc <- NULL
phys_dat2$Rndrng_Prvdr_Cntry <- NULL

### Provider info
phys_dat2$Rndrng_Prvdr_Type <- NULL
phys_dat2 <- mutate(phys_dat2, Rndrng_Prvdr_Mdcr_Prtcptg_Ind = ifelse(Rndrng_Prvdr_Mdcr_Prtcptg_Ind == 'Y', 1, 0))
phys_dat2$Drug_Sprsn_Ind <- NULL # Seems irrelevant
phys_dat2$Med_Sprsn_Ind <- NULL # See above

### Dealing with data suppression
  # I'm going to just replace the suppressed data with 5 (counts lower than 11 are suppressed, 5 is middle of 0 and 10)
  # https://freerangestats.info/blog/2018/11/06/suppressed-data
phys_dat2$Tot_Benes[is.na(phys_dat$Tot_Benes)] <- 5
phys_dat2$Bene_CC_BH_ADHD_OthCD_V1_Pct[is.na(phys_dat$Bene_CC_BH_ADHD_OthCD_V1_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_BH_ADHD_OthCD_V1_Pct)] # This is a percentage of total beneficiaries, which is suppressed if count is less than 11
phys_dat2$Bene_CC_BH_Tobacco_V1_Pct[is.na(phys_dat$Bene_CC_BH_Tobacco_V1_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_BH_Tobacco_V1_Pct)]
phys_dat2$Bene_CC_BH_Alz_NonAlzdem_V2_Pct[is.na(phys_dat$Bene_CC_BH_Alz_NonAlzdem_V2_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_BH_Alz_NonAlzdem_V2_Pct)]
phys_dat2$Bene_CC_BH_Anxiety_V1_Pct[is.na(phys_dat$Bene_CC_BH_Anxiety_V1_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_BH_Anxiety_V1_Pct)]
phys_dat2$Bene_CC_BH_Bipolar_V1_Pct[is.na(phys_dat$Bene_CC_BH_Bipolar_V1_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_BH_Bipolar_V1_Pct)]
phys_dat2$Bene_CC_BH_Mood_V2_Pct[is.na(phys_dat$Bene_CC_BH_Mood_V2_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_BH_Mood_V2_Pct)]
phys_dat2$Bene_CC_BH_Depress_V1_Pct[is.na(phys_dat$Bene_CC_BH_Depress_V1_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_BH_Depress_V1_Pct)]
phys_dat2$Bene_CC_BH_PD_V1_Pct[is.na(phys_dat$Bene_CC_BH_PD_V1_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_BH_PD_V1_Pct)]
phys_dat2$Bene_CC_BH_PTSD_V1_Pct[is.na(phys_dat$Bene_CC_BH_PTSD_V1_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_BH_PTSD_V1_Pct)]
phys_dat2$Bene_CC_BH_Schizo_OthPsy_V1_Pct[is.na(phys_dat$Bene_CC_BH_Schizo_OthPsy_V1_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_BH_Schizo_OthPsy_V1_Pct)]
phys_dat2$Bene_CC_PH_Asthma_V2_Pct[is.na(phys_dat$Bene_CC_PH_Asthma_V2_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_PH_Asthma_V2_Pct)]
phys_dat2$Bene_CC_PH_Afib_V2_Pct[is.na(phys_dat$Bene_CC_PH_Afib_V2_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_PH_Afib_V2_Pct)]
phys_dat2$Bene_CC_PH_Cancer6_V2_Pct[is.na(phys_dat$Bene_CC_PH_Cancer6_V2_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_PH_Cancer6_V2_Pct)]
phys_dat2$Bene_CC_PH_CKD_V2_Pct[is.na(phys_dat$Bene_CC_PH_CKD_V2_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_PH_CKD_V2_Pct)]
phys_dat2$Bene_CC_PH_COPD_V2_Pct[is.na(phys_dat$Bene_CC_PH_COPD_V2_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_PH_COPD_V2_Pct)]
phys_dat2$Bene_CC_PH_Diabetes_V2_Pct[is.na(phys_dat$Bene_CC_PH_Diabetes_V2_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_PH_Diabetes_V2_Pct)]
phys_dat2$Bene_CC_PH_HF_NonIHD_V2_Pct[is.na(phys_dat$Bene_CC_PH_HF_NonIHD_V2_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_PH_HF_NonIHD_V2_Pct)]
phys_dat2$Bene_CC_PH_Hyperlipidemia_V2_Pct[is.na(phys_dat$Bene_CC_PH_Hyperlipidemia_V2_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_PH_Hyperlipidemia_V2_Pct)]
phys_dat2$Bene_CC_PH_Hypertension_V2_Pct[is.na(phys_dat$Bene_CC_PH_Hypertension_V2_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_PH_Hypertension_V2_Pct)]
phys_dat2$Bene_CC_PH_IschemicHeart_V2_Pct[is.na(phys_dat$Bene_CC_PH_IschemicHeart_V2_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_PH_IschemicHeart_V2_Pct)]
phys_dat2$Bene_CC_PH_Osteoporosis_V2_Pct[is.na(phys_dat$Bene_CC_PH_Osteoporosis_V2_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_PH_Osteoporosis_V2_Pct)]
phys_dat2$Bene_CC_PH_Parkinson_V2_Pct[is.na(phys_dat$Bene_CC_PH_Parkinson_V2_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_PH_Parkinson_V2_Pct)]
phys_dat2$Bene_CC_PH_Arthritis_V2_Pct[is.na(phys_dat$Bene_CC_PH_Arthritis_V2_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_PH_Arthritis_V2_Pct)]
phys_dat2$Bene_CC_PH_Stroke_TIA_V2_Pct[is.na(phys_dat$Bene_CC_PH_Stroke_TIA_V2_Pct)] <- 
  5 / phys_dat2$Tot_Benes[is.na(phys_dat$Bene_CC_PH_Stroke_TIA_V2_Pct)]

phys_dat <- phys_dat2

## tax_dat

### Convert variables to numeric
tax_dat_STATE <- tax_dat$STATE
tax_dat <- as.data.frame(lapply(tax_dat, function(x) as.numeric(gsub(',', '', x))))
tax_dat$STATE <- tax_dat_STATE

### Let's merge the income brackets for each state

tax_dat2 <- tax_dat

  # https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/reshape
tax_dat3 <- reshape(
  tax_dat2,
  idvar = 'STATE',
  timevar = 'AGI_STUB',
  direction = 'wide'
)
tax_dat <- tax_dat3


# Remove unnecessary objects

rm(phys_dat2, tax_dat_STATE, tax_dat2, tax_dat3)


# Display cleaned dataframes
head(phys_dat)
head(tax_dat)
```

```{r}
# Check for remaining NAs in phys_dat

sapply(phys_dat, function(x) sum(is.na(x)))
```

```{r}
# Except for RUCA, The remaining NAs have no suppression method specified.  Let's remove them.

phys_dat2 <- phys_dat
phys_dat2$Drug_Tot_HCPCS_Cds <- NULL
phys_dat2$Drug_Tot_Benes <- NULL
phys_dat2$Drug_Tot_Srvcs <- NULL
phys_dat2$Drug_Sbmtd_Chrg <- NULL
phys_dat2$Drug_Mdcr_Alowd_Amt <- NULL
phys_dat2$Drug_Mdcr_Pymt_Amt <- NULL
phys_dat2$Drug_Mdcr_Stdzd_Amt <- NULL
phys_dat2$Med_Tot_HCPCS_Cds <- NULL
phys_dat2$Med_Tot_Benes <- NULL
phys_dat2$Med_Tot_Srvcs <- NULL
phys_dat2$Med_Sbmtd_Chrg <- NULL
phys_dat2$Med_Mdcr_Alowd_Amt <- NULL
phys_dat2$Med_Mdcr_Pymt_Amt <- NULL
phys_dat2$Med_Mdcr_Stdzd_Amt <- NULL
phys_dat2$Bene_Age_LT_65_Cnt <- NULL
phys_dat2$Bene_Age_65_74_Cnt <- NULL
phys_dat2$Bene_Age_75_84_Cnt <- NULL
phys_dat2$Bene_Age_GT_84_Cnt <- NULL
phys_dat2$Bene_Feml_Cnt <- NULL
phys_dat2$Bene_Male_Cnt <- NULL
phys_dat2$Bene_Race_Wht_Cnt <- NULL
phys_dat2$Bene_Race_Black_Cnt <- NULL
phys_dat2$Bene_Race_API_Cnt <- NULL
phys_dat2$Bene_Race_Hspnc_Cnt <- NULL
phys_dat2$Bene_Race_NatInd_Cnt <- NULL
phys_dat2$Bene_Race_Othr_Cnt <- NULL
phys_dat2$Bene_Dual_Cnt <- NULL


# Remove rows with unknown RUCA
phys_dat2 <- phys_dat2[!is.na(phys_dat2$Rndrng_Prvdr_RUCA), ]


sapply(phys_dat2, function(x) sum(is.na(x)))
phys_dat <- phys_dat2
rm(phys_dat2)
```


### First let's see how medicare payments vary by state

Let's see which states are significant predictors in linear regression

```{r}
phys_dat_states_only <- phys_dat[c('Rndrng_Prvdr_State_Abrvtn', 'Tot_Mdcr_Stdzd_Amt')]
lm_states <- lm(Tot_Mdcr_Stdzd_Amt ~ Rndrng_Prvdr_State_Abrvtn, phys_dat_states_only)
summary(lm_states)
```

AL, AR, AZ, CA, DE, FL, IL, KS, MD, MS, NJ, NV, OK, SC, TN, TX, VA are significant predictors at a significance level of 0.05.

### How do the stats of different state's taxes influence medicare payments?

```{r}
# First let's merge the datasets

mdcr_taxes_dat <- merge(phys_dat[c('Rndrng_Prvdr_State_Abrvtn', 'Tot_Mdcr_Stdzd_Amt')], tax_dat, by.x = 'Rndrng_Prvdr_State_Abrvtn', by.y = 'STATE')
```

With over 1000 predictors and 100000 observations it is unrealistic to test each one.

Let's instead try using stepwise selection.  

```{r}
#set.seed(1)
# Reduce dataset with stratified sampling
mdcr_taxes_samp <- mdcr_taxes_dat %>%
  group_by(Rndrng_Prvdr_State_Abrvtn) %>%
  sample_n(50)

# https://cran.r-project.org/web/packages/olsrr/index.html
# https://cran.r-project.org/web/packages/olsrr/vignettes/variable_selection.html
library(olsrr)
lm_taxes <- lm(Tot_Mdcr_Stdzd_Amt ~ ., mdcr_taxes_samp[, 2:dim(mdcr_taxes_dat)[2]]) # Do not use the state abbreviation as predictor
ols_step_forward_p(lm_taxes)
```

Forward stepwise selection using AIC from the reduced sample estimates a linear regression model using A00700.4 (state and local income tax refunds amount for AGI stub 4), N00700.2 (number of returns with state and local income tax refunds for AGI stub 2), A85300.1 (net investment income tax for AGI stub 1), MVITA.0 (number of military volunteer prepared returns for AGI stub 0), A18800.8 (personal property taxes amount for AGI stub 8), A11520.9 (refundable child and dependent care credit for AGI stub 9), A19570.2 (investment interest paid amount for AGI stub 2), A19500.9 (home mortgage from personal seller amount for AGI stub 9), A11450.9 (qualified sick and family leave credit for leave taken before april 1 2021 amount for AGI stub 9), A11530.10 (qualified sick and family leave credit for leave taken after march 31 2021 for AGI stub 10), and A85530.2 (additional medicare tax for AGI stub 2).

It should be noted that this model has a very low r^2 value and poor predictive power.  A85300.1, A11530.10, and A85530.2 are not significant.

```{r}
lm_taxes_only <- lm(Tot_Mdcr_Stdzd_Amt ~ A00700.4 + N00700.2 + A85300.1 + MVITA.0 + A18800.8 + A11520.9 + A19570.2 + A19500.9 + A11530.10 + A85530.2, mdcr_taxes_dat)
summary(lm_taxes_only)
```

### Are these predictors still significant when utilized alongside variables from phys_dat?

```{r}
# Let's determine the best predictors from phys_dat
phys_dat$Tot_Mdcr_Pymt_Amt <- NULL
phys_samp <- phys_dat %>%
  group_by(Rndrng_Prvdr_State_Abrvtn) %>%
  sample_n(50)
lm_phys <- lm(Tot_Mdcr_Stdzd_Amt ~ . - Rndrng_NPI - Rndrng_Prvdr_State_Abrvtn - Rndrng_Prvdr_Mdcr_Prtcptg_Ind + factor(Rndrng_Prvdr_Mdcr_Prtcptg_Ind), phys_samp)
ols_step_forward_p(lm_phys)
```

Tot_Mdcr_Alowd_Amt, Bene_Ndual_Cnt, Tot_Benes, Tot_Srvcs, Tot_Sbmtd_Chrg, Bene_CC_BH_Mood_V2_Pct, Bene_CC_PH_Afib_V2_Pct, Bene_CC_BH_Ptsd_V1_Pct, Tot_HCPCS_Cds, Bene_CC_PH_Parkinson_V2_Pct, Bene_CC_PH_Hyperlipidemia_V2_Pct, Bene_Avg_Risk_Scre, Rndrng_Prvdr_RUCA, and Bene_CC_PH_COPD_V2_Pct

```{r}
lm_phys_final <- lm(Tot_Mdcr_Stdzd_Amt ~ Tot_Mdcr_Alowd_Amt + Bene_Ndual_Cnt + Tot_Benes + Tot_Srvcs +  Tot_Sbmtd_Chrg+ Bene_CC_BH_Mood_V2_Pct+ Bene_CC_PH_Afib_V2_Pct + Bene_CC_BH_PTSD_V1_Pct + Tot_HCPCS_Cds+ Bene_CC_PH_Parkinson_V2_Pct+ Bene_CC_PH_Hyperlipidemia_V2_Pct+ Bene_Avg_Risk_Scre+ Rndrng_Prvdr_RUCA+ Bene_CC_PH_COPD_V2_Pct, phys_dat)
summary(lm_phys_final)
```

```{r}
# Full model
final_dat <- merge(phys_dat[c('Rndrng_Prvdr_State_Abrvtn','Tot_Mdcr_Alowd_Amt',  'Tot_Mdcr_Stdzd_Amt','Bene_Ndual_Cnt','Tot_Benes','Tot_Srvcs','Tot_Sbmtd_Chrg', 'Bene_CC_BH_Mood_V2_Pct', 'Bene_CC_PH_Afib_V2_Pct', 'Bene_CC_BH_PTSD_V1_Pct', 'Tot_HCPCS_Cds', 'Bene_CC_PH_Parkinson_V2_Pct', 'Bene_CC_PH_Hyperlipidemia_V2_Pct', 'Bene_Avg_Risk_Scre','Rndrng_Prvdr_RUCA', 'Bene_CC_PH_COPD_V2_Pct')], tax_dat[c('A00700.4', 'N00700.2', 'A85300.1', 'A18800.8', 'A11520.9', 'A19570.2', 'A19500.9', 'A11530.10', 'STATE')], by.x = 'Rndrng_Prvdr_State_Abrvtn', by.y = 'STATE')
lm_final <- lm(Tot_Mdcr_Stdzd_Amt ~ . - Rndrng_Prvdr_State_Abrvtn, final_dat)
summary(lm_final)
```

