---
title: "STATS506 Problem Set #1"
author: "Bryan Li"
format: html
editor: visual
---

## Problem #1 - Wine Data

### a.

```{r}
wine <- read.table(file = "Datasets/wine/wine.data",
                   sep = ',',
                   col.names = c('class', 'alcohol', 'malic_acid', 'ash', 'alcalinity_of_ash', 'magnesium', 'total_phenols', 'flavanoids', 'nonflavanoid_phenols', 'proanthocyanins', 'color_intensity', 'hue', 'OD280_and_OD314_of_diluted_wines', 'proline'))
head(wine)
```

### b.

```{r}
table(wine$class)
```

wine.names states that there are 59 wines of class 1, 71 of class 2, and 48 of class 3. This is confirmed by the previous table.

### c.

1.  

```{r}
cor(wine$alcohol, wine$color_intensity)
```

2.  

```{r}
cor(wine$alcohol[wine$class == 1], wine$color_intensity[wine$class == 1])
cor(wine$alcohol[wine$class == 2], wine$color_intensity[wine$class == 2])
cor(wine$alcohol[wine$class == 3], wine$color_intensity[wine$class == 3])
```

Class 1 has the highest, class 2 the lowest.

3.  

```{r}
wine$alcohol[wine$color_intensity == max(wine$color_intensity)]
```

4.  

```{r}
proanthocyanins_greater_ash_vec <- c(wine$proanthocyanins > wine$ash)
proanthocyanins_greater_ash_perc <- mean(proanthocyanins_greater_ash_vec)
print(paste("Percentage of Wines with a Higher Content of Proanthocyanins than Ash:", proanthocyanins_greater_ash_perc, sep = ' '))
```

### d.

```{r}
wine_mean_table <- data.frame(Class = c('1', '2', '3', 'Overall'))
for (i in 2:ncol(wine)) {
  class_means <- tapply(wine[, i], wine$class, mean)
  overall_mean <- mean(wine[, i])
  wine_mean_table[names(wine)[i]] <- c(class_means[1], class_means[2], class_means[3], overall_mean)
}
wine_mean_table
```

### e.

```{r}
test_stat <- function(mu1, mu2, s1, s2, n1, n2) {
  return((mu1 - mu2) / sqrt( ((s1^2)/n1) + ((s2^2)/n2)))
}

dof <- function(s1, s2, n1, n2) {
  numer <- ( ((s1^2)/n1) + ((s2^2)/n2) )^2
  denom <- ( ( (((s1^2)/n1)^2 ) / (n1 - 1) ) + ( (((s2^2)/n2)^2 ) / (n2 - 1)) )
  return(numer/denom)
}

t_test_diff_2_tail <- function(mu1, mu2, s1, s2, n1, n2) {
  test_stat <- test_stat(mu1, mu2, s1, s2, n1, n2)
  dof <- dof(s1, s2, n1, n2)
  
  if (test_stat > 0.5) {
    return(pt(test_stat, dof, lower.tail = FALSE) * 2)
  } else {
    return(pt(test_stat, dof, lower.tail = TRUE) * 2)
  }
}
```

```{r}
mean1 <- mean(wine$total_phenols[wine$class == 1])
sd1 <- sd(wine$total_phenols[wine$class == 1])
n1 <- length(wine$total_phenols[wine$class == 1])

mean2 <- mean(wine$total_phenols[wine$class == 2])
sd2 <- sd(wine$total_phenols[wine$class == 2])
n2 <- length(wine$total_phenols[wine$class == 2])

mean3 <- mean(wine$total_phenols[wine$class == 3])
sd3 <- sd(wine$total_phenols[wine$class == 3])
n3 <- length(wine$total_phenols[wine$class == 3])


# Class 1 vs 2
# H0: mu1 = mu2
# H1: mu1 != mu2
print(paste('P-Value of Class 1 vs 2:', t_test_diff_2_tail(mean1, mean2, sd1, sd2, n1, n2), sep = ''))
print(paste('Class 1 vs 2 Test Statistic:', test_stat(mean1, mean2, sd1, sd2, n1, n2), sep = ''))
print(paste('Class 1 vs 2 DoF:', dof(sd1, sd2, n1, n2), sep = ''))

# Class 2 vs 3
# H0: mu2 = mu3
# h1: mu2 != mu3
print(paste('P-Value of Class 2 vs 3:', t_test_diff_2_tail(mean2, mean3, sd2, sd3, n2, n3), sep = ''))
print(paste('Class 2 vs 3 Test Statistic:', test_stat(mean2, mean3, sd2, sd3, n2, n3), sep = ''))
print(paste('Class 2 vs 3 DoF:', dof(sd2, sd3, n2, n3), sep = ''))

# Class 1 vs 3
# H0: mu1 = mu3
# H1: mu1 != mu3
print(paste('P-Value of Class 1 vs 3:', t_test_diff_2_tail(mean1, mean3, sd1, sd3, n1, n3), sep = ''))
print(paste('Class 1 vs 3 Test Statistic:', test_stat(mean1, mean3, sd1, sd3, n1, n3), sep = ''))
print(paste('Class 1 vs 3 DoF:', dof(sd1, sd3, n1, n3), sep = ''))
```

All three tests are statistically significant at a significance value of 0.05.  At $\alpha = 0.05$ we reject the null hypothesis for all three tests, the mean total phenols are different between all classes of wine.

## Problem #2 - AskAManager.org Data

### a.
